Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2023-04-05 StatPaper/APA-lato_New/APA/wileyNJD-APA.tex
File Changes

\abstract[Summary]{Neural network-based survival methods can model data-driven
covariate interactions. While these methods have led to better
predictive performance than regression-based approaches, they cannot
model both time-varying interactions and complex baseline hazards. To
address this, we propose Case-Base Neural Networks (CBNN) as a new
approach that combines the case-base sampling framework with flexible
architectures. Our method naturally accounts for censoring and does not
require method specific hyperparameters. Using a novel sampling scheme
and data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. We highlight the benefit of combining
case-base sampling with deep learning to provide a simple and flexible
modeling framework for data-driven, time-varying interaction modeling of
survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

Neural networks provide a data-driven approach to approximating
interaction terms. For example, DeepSurv is a neural network-based
proportional hazards model that implements the Cox partial
log-likelihood as a custom loss function \citep{katzman2018DeepSurv},
resulting in a stepwise absolute risk curve that cannot accommodate
time-varying interactions. Compared to Cox regression, DeepSurv shows
better performance on the Study to Understand Prognoses Preferences and
Risk Treatments (SUPPORT) dataset \citep{knaus1995SUPPORT}. To handle
non-proportional hazards, a modification to the loss function was
proposed \citep{faraggi1995neural}. To remove the need for tuning the
number of layers and nodes, the concept of extreme learning machines has
been applied to a Cox neural network model \citep{wang2019extreme}.

As an alternative method that assumes a baseline hazard distribution,
DeepHit specifies each survival time of interest in the model and
directly estimates survival curves, rather than deriving a hazard
function \citep{lee2018DeepHit}. It assumes an inverse Gaussian
distribution as the baseline hazard and it outperforms DeepSurv on the
Molecular Taxonomy of Breast Cancer International Consortium (METABRIC)
dataset \citep{curtis2012genomic}.

Providing further flexibility, Deep Survival Machines (DSM) is a
parametric survival model using neural networks with a mixture of
distributions as the baseline hazard \citep{dsmPaper}. On both the
SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

@@

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter decisions can affect the
estimated parameters and were chosen during a set of initial simulations
to determine if CBNN can learn complex interactions in practice. We set
a max number of epochs to be \(2000\), batch size as \(512\), learning
rate as \(10^{-3}\), decay as \(10^{-7}\), patience as \(10\) epochs,
\(\{50,50,25,25\}\) nodes in each hidden layer with \(50\%\) dropout at
each layer, a minimum delta loss on the validation set of \(10^{-7}\)
over 10 epochs and Adam \citep{gulli2017} as our optimizer. These
choices may permit the model to approximate higher-order interactions
while preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test) for each
method. We select the best weights after training based on validation
loss \citep{gulli2017}.

@@

\hypertarget{sims}{%
\section{Simulation studies}\label{sims}}

In this section, we use simulated data to evaluate the performance of
CBNN and compare our approach with existing regression-based (Cox, CBLR)
and neural network-based (DeepHit, DeepSurv, DSM) methods. We specify a
linear combination of each covariate as the linear predictor in
regression-based approaches (Cox, CBLR), which contrasts with neural
network approaches that allow for non-linear interactions. We simulate
data under a simple exponential model and a complex baseline hazard with
time-varying interactions, each with 10\% random censoring. For both
settings, we simulate three covariates:

\[
z_{1} \sim \textrm{Bernoulli}(0.5) \qquad \qquad
z_{2} \sim \begin{cases}
 N(0,0.5) & \textrm{if } z_{1}=0\\ 
 N(10,0.5) & \textrm{if } z_{1}=1
\end{cases} \qquad \qquad
z_{3} \sim \begin{cases}
 N(8,0.5) & \textrm{if } z_{1}=0\\ 
 N(-3,0.5) & \textrm{if } z_{1}=1.
\end{cases}
\]

The DeepHit-specific hyperparameter alpha is set to \(0.5\) (equal
weight between its negative log-likelihood and ranking losses
\citep{lee2018DeepHit}). We modify the \textbf{DeepSurvivalMachines}
\citep{dsmPaper} package to include dropout and a minimum delta loss
during the fitting process. For DSM, we define a mixture model of six
Weibull distributions for the baseline hazard. All other hyperparameters
are held constant across all neural network methods in both the
simulation studies and real data applications.

Besides the methods mentioned above, we include the Optimal model in our
comparisons using CBLR. That is, we include the exact functional form of
the covariates in a CBLR model (referred to as Optimal for simplicity).
We calculate \(t\)-based 95\% confidence intervals using 100
replications of the simulated data. For all analyses, we use 80\% for
training and 20\% for the test set. 20\% of the training set is kept for
validation at each epoch. We predict risk functions \(F\) using
\eqref{eqn:ci2} for individuals in the test set, which are used to
calculate our \(C_{IPCW}\) and IPA scores.

\hypertarget{performance-metrics}{%
\subsection{Performance metrics}\label{performance-metrics}}

We use two metrics to assess the performance of the different methods of
interest: 1) (IPA) \citep{kattan2018index} and 2) inverse probability
censoring weights-adjusted concordance index (\(C_{IPCW}\))
\citep{uno2011}, which we define below. Both these time-dependent
metrics provide transparency as to when in follow-up time each model may
perform better than the others.

@@

@@

\hypertarget{simple-simulation-constant-baseline-hazard}{%
\subsection{Simple simulation: constant baseline
hazard}\label{simple-simulation-constant-baseline-hazard}}

We simulate data from a simple model that primarily depends on a
constant baseline hazard:
\[\log h(t \mid X_i) = \beta_{{1}}z_{1}+\beta_{{2}}z_{2}+\beta_{{3}}z_{3},\]
The covariate effects are given by
\(\beta_{{1}}=0.1, \beta_{{2}}=0.1, \beta_{{3}}=0.1\). Once we simulate
survival times, we introduce 10\% random censoring.

\hypertarget{performance-comparison-in-simple-simulation}{%
\subsubsection{Performance comparison in simple
simulation}\label{performance-comparison-in-simple-simulation}}

Figure \ref{fig:megaPlot} A, B and Table \ref{tab:megaTable} A show the
results for the simple simulation. The regression-based methods (CBLR,
Cox, Optimal) outperform the neural network ones in the simple
simulation setting. Among the neural network approaches, CBNN
outperforms all other methods in terms of both IPA and \(C_{IPCW}\)
(Figure \ref{fig:megaPlot} A, B). Specifically, we see CBNN is
consistent across time with smaller confidence intervals compared to
DeepHit, DeepSurv and DSM.

In this simple setting, the regression models are much closer to the
Optimal model, while the neural network models perform worse than the KM
null model. The wide confidence bands in \(C_{IPCW}\) suggest the neural
network models may be over-parameterized.

@@

This simulation demonstrates performance with the presence of a complex
baseline hazard and a time-varying interaction. Originally used to show
the spline-based hazard model proposed by Royston and Parmar
\citep{royston2002flexible}, the breast cancer dataset provides a
complex hazard from which we simulate, available in the
\textbf{flexsurv} R package \citep{flexsurv}. To increase the complexity
of our data-generating mechanism for this simulation, we design the
model as follows: \begin{align}
\log h(t \mid X_i) =\sum_{i=1}^{5} (\gamma_{i} \cdot \psi_{i}) + \beta_{{1}} (z_{1}) + \beta_{{2}} (z_{2})+ \beta_{{3}} (z_{3})+ \tau_{1} ( z_{1} \cdot z_{2} \cdot t)+ \tau_{2} ( z_{1} \cdot z_{3})+ \tau_{3} (z_{2} \cdot z_{3}), \nonumber
\end{align}

\begin{table}
\caption{Four tables representing performance at certain follow-up times for the simple simulation, complex simulation, SUPPORT and METABRIC. Each table shows performance for each method in each study at $25\%$, $50\%$. $75\%$ and $100\%$ of follow-up time. The bold elements show the best model for each study, at each follow-up time of interest. These tables are included to provide exact measures at certain intervals. The models of interest are: Cox, case-base with logistic regression (CBLR), DeepSurv, DeepHit, Case-Base Neural Network (CBNN), Optimal and Deep Survival Machines (DSM).}
\label{tab:megaTable}

@@

where
\(\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2\)
and \(\psi\) are basis splines. The \(gamma\) coefficients are obtained
from an intercept-only cubic splines model with three knots using the
\emph{flexsurvspline} function from the \textbf{flexsurv} package
\citep{flexsurv}. Note that we fix these values for the analysis. The
\(\beta\) coefficients represent direct effects, \(\tau_{2}\) and
\({\tau_3}\) represent interactions and \(\tau_{1}\) is a time-varying
interaction.

@@

Figure \ref{fig:megaPlot} C, D and Table \ref{tab:megaTable} B show the
performance over time on a test set in the complex simulation. Apart
from the Optimal regression model, CBNN outperforms the competing models
when examining IPA and up to the 75-percentile of follow-up time for
\(C_{IPCW}\). We expected the Optimal model to perform best in both
metrics. However, this was not the case for \(C_{IPCW}\) and may be due
to an artifact of concordance-based metrics, where a misspecified model
may perform better than a correctly specified one
\citep{cindexfails2019}. We attribute the performance of CBNN to its
flexibility in modeling time-varying interactions and baseline hazard,
flexibility the other neural network models do not have.

@@

\hypertarget{casestudies}{%
\section{Application to SUPPORT and METABRIC data}\label{casestudies}}

Our complex simulation demonstrates the superior performance of CBNN in
ideal conditions with clean data. To obtain a more realistic performance
assessment, we compared models using two real datasets with a
time-to-event outcome. The first case study examines the SUPPORT dataset
\citep{knaus1995SUPPORT}. The second case study examines the METABRIC
dataset \citep{curtis2012genomic}. We use the same hyperparameters as in
the simulation studies. As we do not know the true model for the real
data, we exclude the Optimal model. We split the datasets keeping 20\%
of the observations as a test set. 20\% of the training set is kept
aside for validation at each epoch. We predict risk functions for
everyone in the test set, which is used to calculate our metrics. We
conduct 100 bootstrap re-samples for the real data applications to
obtain confidence intervals.

\hypertarget{performance-evaluation-using-the-support-dataset}{%
\subsection{Performance evaluation using the SUPPORT
dataset}\label{performance-evaluation-using-the-support-dataset}}

The SUPPORT dataset tracks the time until death for seriously ill
patients at five American hospitals \citep{knaus1995SUPPORT}. We use a
pre-processed version of the dataset made available in the DeepSurv
package \citep{katzman2018DeepSurv}. This dataset contains 9104 samples
and 14 covariates (age, sex, race, number of comorbidities, presence of
diabetes, presence of dementia, presence of cancer, mean arterial blood
pressure, heart rate, respiration rate, temperature, white blood cell
count, serum's sodium and serum's creatinine)
\citep{katzman2018DeepSurv}. Patients with missing features were
excluded and 68.10\% of the patients died during the 5.56-year study
period \citep{katzman2018DeepSurv}.

Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C
demonstrates the performance over time on a test set. The regression
models (CBLR, Cox) perform best considering IPA, followed by CBNN from
the \(25^{th}\) to \(100^{th}\) percentile of follow-up time. We note a
drop in performance for CBNN before the \(25^{th}\) percentile of
follow-up. For \(C_{IPCW}\), CBNN outperforms the competing models
consistently over follow-up time. Note that performance is similar for
all models, aside from DeepSurv whose \(C_{IPCW}\) is lower than the
rest (Figure \ref{fig:megaPlot} E, F and Table \ref{tab:megaTable} C).

\hypertarget{performance-evaluation-using-the-metabric-dataset}{%
\subsection{Performance evaluation using the METABRIC
dataset}\label{performance-evaluation-using-the-metabric-dataset}}

METABRIC is a 30-yearlong study aiming to discover the molecular drivers
of breast tumors, following 2000 individuals with breast cancer until
death \citep{curtis2012genomic}. They described these growths as primary
invasive breast carcinomas, with the goal of discovering both genetic
and clinical risk factors for breast cancer survival
\citep{curtis2012genomic}. We used the processed dataset made available
through DeepSurv \citep{katzman2018DeepSurv}, which includes 1980
patients of which 57.72\% die due to breast cancer within a median 10
years of follow-up \citep{katzman2018DeepSurv}. There are 10 covariates
in total: time, 4 RNA-Seq gene expressions (MKI67, EGFR, PGR and ERBB2)
and 5 clinical features (hormone treatment indicator, radiotherapy
indicator, chemotherapy indicator, ER-positive indicator and age at
diagnosis) \citep{katzman2018DeepSurv}.

Figure \ref{fig:megaPlot} G, H and Table \ref{tab:megaTable} D shows the
performance on a test set over time. The IPA scores suggest that
regression models outperform competing models on this dataset, as all
the neural network models are equal to or perform worse than KM over
follow-up time. Our CBNN model is comparable to KM until around the
50-percentile of follow-up time, after which CBNN, DeepSurv and DSM drop
in performance. The \(C_{IPCW}\) produces a different ranking. Our CBNN
model outperforms the other models up to the \(25^{th}\) percentile of
follow-up time, whereas DeepHit performs best from the \(50^{th}\) to
\(75^{th}\) percentile of follow-up. With this metric, CBNN and DeepHit
outperform the regression models. This disagreement between IPA and
\(C_{IPCW}\) may be due to the misspecification issue of
concordance-based metrics \citep{cindexfails2019}. The neural network
models may be over-parameterized, as shown by the wide confidence bands
in \(C_{IPCW}\).

@@

CBNN models survival outcomes by using neural networks on case-base
sampled data. We incorporate follow-up time as a feature, providing a
data-driven estimate of a flexible baseline hazard and time-varying
interactions in our hazard function. The three competing neural network
models we evaluated cannot model time-varying interactions by design
\citep{dsmPaper} \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}. DSM
requires a mixture component distributions for a flexible baseline
hazard to be fit \citep{dsmPaper}. As our goal is to fix the design and
compare performance, we did not change any shared hyperparameters. With
our choice of shared hyperparameters and model design, DSM did not
converge in the complex simulation and SUPPORT case study. Despite this
limitation, we include this method when it did function as it can fit
flexible baseline hazards. Compared to CBNN, the remaining two models
also have limitations. DeepSurv is a proportional hazards model and does
not estimate the baseline hazard \citep{katzman2018DeepSurv}. DeepHit
requires an alpha hyperparameter, is restricted to a single distribution
for the baseline hazard and models the survival function directly
\citep{lee2018DeepHit}. The alternative neural network methods match on
time, while CBNN models time directly.

To assess performance among these models, we use both IPA and
\(C_{IPCW}\) metrics. Concordance-based measures are commonly used in
survival analysis to compare models and we opt to keep them in our
analyses. However, \(C_{IPCW}\) is a non-proper metric and may cause
misspecified models to appear better than they should
\citep{cindexfails2019}. Therefore, we contextualize our \(C_{IPCW}\)
results in relation to IPA, a proper scoring rule. The model rankings
between \(C_{IPCW}\) and IPA differed for both the complex simulation
and METABRIC application. Wider \(C_{IPCW}\) confidence intervals for
DeepHit and DeepSurv show potential misspecification of the models
(Figure \ref{fig:megaPlot} D, H).

With this interpretation of \(C_{IPCW}\) in mind, we assess two
simulations with minimal noise and two case studies in real scenarios.
The simple simulation demonstrates potential pitfalls associated with
neural network models, particularly overparameterization. All neural
network-based approaches performed worse than null KM model (IPA score),
while the regression-based performed better. We attribute this to
potential overparameterization in the neural network models, as the wide
confidence intervals suggest over-fitting, even with dropout. Both
DeepSurv and DSM are affected, while DeepHit and CBNN are less so. This
is a limitation to our strategy for evaluating the methods, which uses a
fixed study design across all assessments.

From this baseline benchmark, we move on to a complex simulation that
requires a method that can learn both time-varying interactions and have
a flexible baseline hazard. Here, CBNN demonstrates a distinct advantage
over all other methods. The regression models show improved performance
over the null KM model, while the competing neural network models
performed worse. Based on our complex simulation results (Figure
\ref{fig:megaPlot} C, D and Table \ref{tab:megaTable} B), CBNN
outperforms the competitors when time-varying interactions and a complex
baseline hazard are present. This simulation shows how CBNN can perform
under ideal conditions, while the following two analyses on real data
serve to assess its performance in realistic conditions.

In the SUPPORT and METABRIC case studies, flexibility in both
interaction modeling and baseline hazard did not improve CBNNs relative
performance, suggesting that this flexibility does not aid prediction in
either case study. From a biological perspective, the 30-year follow-up
time in the METABRIC study may contain competing causes of death. The
causes at the start of the study may not match the causes towards the
end, potentially explaining the drop in performance as we reach later
survival times with fewer individuals. The baseline hazard is unlikely
to cause this drop as DSM is the most flexible competing model in our
comparison. Over-fitting is also unlikely given the tight confidence
intervals. Further research is required to determine the cause of the
drop in performance seen in the METABRIC case study. In both case
studies, CBNN outperforms the competing neural network methods.

Taken together, CBNN outperforms all competitors in the complex
simulation, demonstrating its value in survival settings that may
involve time-varying interactions and a complex baseline hazard. Once we
perform case-base sampling and adjust for the sampling bias, we can use
a sigmoid activation function to predict our hazard function. Our
approach simplifies the incorporation of censored individuals, allowing
survival outcomes to be treated as binary ones. Forgoing the requirement
of custom loss functions, CBNN only requires the use of standard
components in machine learning libraries (specifically, the add layer to
adjust for sampling bias and the sigmoid activation function). Due to
the simplicity in its implementation and by extension user experience,
CBNN is both a user-friendly approach to data-driven survival analysis
and is easily extendable to any feed-forward neural network framework.

@@

This is an author contribution text. This is an author contribution text. This is an author contribution text. This is an author contribution text. This is an author contribution text.

@@

\begin{biography}{\includegraphics[width=60pt,height=70pt,draft]{empty}}{\textbf{Author Name.} This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
\end{biography}

>>>>

\abstract[Summary]{Neural network-based survival methods can model data-driven
covariate interactions. While these methods have led to better
predictive performance than regression-based approaches, they cannot
model both time-varying interactions and complex baseline hazards. To
address this, we propose Case-Base Neural Networks (CBNN) as a new
approach that combines the case-base sampling framework with flexible
architectures. Our method naturally accounts for censoring and does not
require method specific hyperparameters. Using a novel sampling scheme
and data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in a simulation and three real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a complex simulation, which highlights the ability of
CBNN to model both a complex baseline hazard and time-varying
interactions, CBNN outperforms all competitors. The first two real data
application shows CBNN outperforming all neural network competitors,
while a second real data application shows competitive performance.
We highlight the benefit of combining case-base sampling with deep
learning to provide a simple and flexible modeling framework for
data-driven, time-varying interaction modeling of
survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

Neural networks provide a data-driven approach to approximating
interaction terms. For example, DeepSurv is a neural network-based
proportional hazards model that implements the Cox partial
log-likelihood as a custom loss function \citep{katzman2018DeepSurv},
resulting in a stepwise absolute risk curve that cannot accommodate
time-varying interactions. Compared to Cox regression, DeepSurv shows
better performance on a real dataset \citep{katzman2018DeepSurv}. To handle
non-proportional hazards, a modification to the loss function was
proposed \citep{faraggi1995neural}.

As an alternative method that assumes a baseline hazard distribution,
DeepHit specifies each survival time of interest in the model and
directly estimates survival curves, rather than deriving a hazard
function . It assumes an inverse Gaussian
distribution as the baseline hazard and it outperformed DeepSurv \citep{lee2018DeepHit}.

Providing further flexibility, Deep Survival Machines (DSM) is a
parametric survival model using neural networks with a mixture of
distributions as the baseline hazard \citep{dsmPaper}. DSM outperformed DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

@@

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter choices can affect the
estimated parameters and are not necessarily the same for each model.
First we fix a test set with 15\% of the data, then we use three-fold cross-validation on the remaining 
data (85\%:15\% training:validation) for a range of hyperparameters (TABLE REF). We track integrated brier 
score on the validation set for each hyperparameter combination, choosing 
the combination with the lowest score for each method. The chosen 
hyperparameters are in (TABLE REF).

@@

\hypertarget{performance-metrics}{%
\section{Performance metrics}\label{performance-metrics}}

To choose our method-specific hyperparameters, we use the Integrated 
Brier Score (IBS) \citep{graf1999}. We use two metrics to assess the 
performance of the different methods of interest on testing data: 1) (IPA) 
\citep{kattan2018index} and 2) inverse probability
censoring weights-adjusted concordance index (\(C_{IPCW}\))
\citep{uno2011}, which we define below. IBS provides a summarized assessment
of performance for each model. (\(C_{IPCW}\)) provides information
about performance up to the survival time of interest. while IPA provide transparency
as to when in follow-up time each model may
perform better than the others.

@@

\hypertarget{integrated-brier-score-ibs}{%
\subsubsection{Integrated Brier Score (IBS)}\label{integrated-brier-score-ibs}}

The IBS is a function of the Brier score (\(BS(t)\)) \citep{graf1999}. as well,
which is defined as \begin{align}
IBS=\int_{t}^{t_{max}}BS(t)dw(t)
\end{align} where $w(t)=\frac{t}{t_{max}}$. As this is a score for a range of follow-up-time,
it is a useful metric to track during cross-validation. The performance at a specific time
is hidden, and is why we opt for the IPA in assessing the final models on a test set.

@@

\hypertarget{sims}{%
\section{Simulation study}\label{sims}}

In this section, we simulate data to evaluate the performance of
CBNN and compare our approach with existing regression-based (Cox, CBLR)
and neural network-based (DeepHit, DeepSurv) methods. We specify a
linear combination of each covariate as the linear predictor in
regression-based approaches (Cox, CBLR), which contrasts with neural
network approaches that allow for non-linear interactions. We simulate
data under a complex baseline hazard with
time-varying interactions, each with 10\% random censoring. For both
settings, we simulate three covariates:

\[
z_{1} \sim \textrm{Bernoulli}(0.5) \qquad \qquad %technically a control and treatement group
z_{2} \sim \begin{cases}
 N(0,0.5) & \textrm{if } z_{1}=0\\ 
 N(1,0.5) & \textrm{if } z_{1}=1
\end{cases} \qquad \qquad
z_{3} \sim N(1,0.5)
\]
Besides the methods mentioned above, we include the Optimal model in our
comparisons using CBLR. That is, we include the exact functional form of
the covariates in a CBLR model (referred to as Optimal for simplicity).
We calculate \(t\)-based 95\% confidence intervals using 100
replications of the simulated data. For all analyses, 15\% of the data is 
kept for testing, while 15\% of the remaining data is kept for validation. 
The remaining data is kept for training. We predict risk functions \(F\) using
\eqref{eqn:ci2} for individuals in the test set, which are used to
calculate our \(C_{IPCW}\) and IPA scores. We conduct 100 bootstrap
re-samples on the training data to obtain confidence intervals.

@@

This simulation demonstrates performance with the presence of a complex
baseline hazard and a time-varying interaction. Inspired by a cancer treatment 
that initially reduces risk of death, but slowly returns as time progresses. 
Originally used to show the spline-based hazard model proposed by Royston and Parmar
\citep{royston2002flexible}, the breast cancer dataset provides a
complex hazard from which we simulate, available in the
\textbf{flexsurv} R package \citep{flexsurv}. To increase the complexity
of our data-generating mechanism for this simulation, we design the
model as follows: \begin{align}
\log h(t \mid X_i) =\sum_{i=1}^{5} (\gamma_{i} \cdot \psi_{i}) + \beta_{{1}} (z_{1}) + \beta_{{2}} (z_{2})+ \beta_{{3}} (z_{3})+ \tau_{1} ( z_{1} \cdot t)+ \tau_{2} (z_{2} \cdot z_{3}), \nonumber
\end{align}

\begin{table}
\caption{Four tables representing performance at certain follow-up times for the simple simulation, complex simulation, SUPPORT and METABRIC. Each table shows performance for each method in each study at $25\%$, $50\%$. $75\%$ and $100\%$ of follow-up time. The bold elements show the best model for each study, at each follow-up time of interest. These tables are included to provide exact measures at certain intervals. The models of interest are: Cox, case-base with logistic regression (CBLR), DeepSurv, DeepHit, Case-Base Neural Network (CBNN) and Optimal}
\label{tab:megaTable}

@@

where
\(\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=-5, \beta_{{2}}=-1, \beta_{{3}}=1, \tau_{1}=0.001, \tau_{2}=-1\)
and \(\psi\) are basis splines. The \(gamma\) coefficients are obtained
from an intercept-only cubic splines model with three knots using the
\emph{flexsurvspline} function from the \textbf{flexsurv} package
\citep{flexsurv}. Note that we fix these values for the analysis. The
\(\beta\) coefficients represent direct effects, \(\tau_{2}\) and
\({\tau_3}\) represent interactions and \(\tau_{1}\) is a time-varying
interaction.

We use a population time plot to visualize incidence density across $z1$ (FIGURE REF).
z1 acts to mimic a treatment and control group. The control group $(z1=0)$ is not impacted
by the time-varying interaction. The treatment group $(z1=1)$ initially acts to protect the group,
while the risk slowly builds with time at the rate of the interaction term $(\tau_{1})$. This creates
a separation between the median time of death between these two groups due to the time-varying
interaction. We expect all competing models (aside from the Optimal model and CBNN) to drop in
performance for the second half of follow-up time.

@@

Figure \ref{fig:megaPlot} A, E and Table \ref{tab:megaTable} A
demonstrates the performance over time on a test set. While examining the IPA score, the optimal
regression model and CBNN perform near identically, leading in performance followed by 
DeepHit, the linear regression models and DeepSurv. We expected the Optimal model to perform best in both
metrics. However, this was not the case for \(C_{IPCW}\) and may be due
to an artifact of concordance-based metrics, where a misspecified model
may perform better than a correctly specified one
\citep{cindexfails2019}. We attribute the performance of CBNN to its
flexibility in modeling time-varying interactions and baseline hazard,
flexibility the other neural network models do not have. We keep this misspecification issue
in mind while interpreting the \(C_{IPCW}\) results in the case studies.

@@

\hypertarget{casestudies}{%
\section{Casestudies}\label{casestudies}}

Our complex simulation demonstrates the superior performance of CBNN in
ideal conditions with clean data. To obtain a more realistic performance
assessment, we compared models using three real datasets with a
time-to-event outcome. The first case study examines multiple myeloma (MM) \citep{myeloma}.
The second examines the relationship between serum free light chain (FLC) and mortality \citep{flc}.
The third examines prostate cancer survival on a publicly available simulation of the Surveillance,
Epidemiology, and End Results (SEER)-medicare prostate cancer data\citep{prostate}.
As we do not know the true model for the real
data, we exclude the Optimal model. After grid search is performed with the same hyperparameter
options, We split the data the same way we did for the simulation (15\% of the data is 
kept for testing, while 15\% of the remaining data is kept for validation and the rest for testing).
We predict risk functions for everyone in the test set, which is used to calculate our metrics. We
conduct 100-fold bootstrap re-samples on the training data to
obtain confidence intervals.

\hypertarget{pe-multiplemyeloma}{%
\subsection{Performance evaluation on multiple myeloma dataset}\label{pe-multiplemyeloma}}
The (MM) dataset tracks 3882 patients seen at the Mayo Clinic from 1947-1996 until death \citep{myeloma}.
We use 2 covariates, year of entry into the study and the time of MM diagnosis\citep{myeloma}. We see 71\% incidence over 23 years\citep{myeloma}.

Figure \ref{fig:megaPlot} B, F and Table \ref{tab:megaTable} B
demonstrates the performance over time on a test set. With the IPA score, CBNN outperforms
The competing models, followed by the linear models, DeepSurv and DeepHit. We note that after
50\% of survival time, all models aside from CBNN perform substantially worse than a null model.
We note with \(C_{IPCW}\), the linear models and DeepSurv perform equivalently, followed by CBNN and
DeepHit.
%paragraph about performance

\hypertarget{pe-flc}{%
\subsection{Performance evaluation on free light chain dataset}\label{pe-flc}}
The FLC dataset is a random sample of half the individuals in $\frac{2}{3}$ of the
residents of Olmsted County over the age of 50\citep{flc}. We have access to 7874 subjects
tracked until death\citep{flc}. We use 5 covariates, total serum FLC (kappa+lambda), age, sex,
serum creatine and monoclonal gammapothy state\citep{flc}. we see 27\% incidence over 14 years\citep{flc}.

Figure \ref{fig:megaPlot} C, G and Table \ref{tab:megaTable} C
demonstrates the performance over time on a test set. With the IPA score, CBNN outperforms
The competing models, followed by the linear models and lastly DeepSurv and DeepHit for the first 60\% of follow-up time. 
After 60\% of follow-up time, DeepHit has similar performance to CBNN. For the \(C_{IPCW}\) score, CBNN outperforms all models,
followed by the linear models and finally DeepSurv. Note that for earlier ranges of survival time, DeepHit is the worst performing model. 
After 25\% of follow-up time, DeepHit is in second place. In the FLC dataset, CBNN has a consistent top ranking for both IPA and  \(C_{IPCW}\).

\hypertarget{pe-prostate}{%
\subsection{Performance evaluation on prostate cancer dataset}\label{pe-prostate}}
The prostate cancer dataset has a record of competing risks\citep{prostate}. As we are only interested in the
single event scenario, we only keep individuals with prostate cancer death or censoring\citep{prostate}.
This subset tracks 11054 individuals with three covariates, differentiation grade, age group and cancer state\citep{prostate}.
There is a 7\% incidence over 10 years.

Figure \ref{fig:megaPlot} C, G and Table \ref{tab:megaTable} C shows the
performance over time on a test set. With the IPA score, the linear models
outperform the neural network ones, followed by CBNN and DeepHit, and
finally DeepSurv. For the \(C_{IPCW}\) score, the linear models, CBNN, DeepHit
perform near identically, side from DeepHit with an initial drop in performance.
DeepSurv performed substantially worse than the other models.

@@

CBNN models survival outcomes by using neural networks on case-base
sampled data. We incorporate follow-up time as a feature, providing a
data-driven estimate of a flexible baseline hazard and time-varying
interactions in our hazard function. The two competing neural network
models we evaluated cannot model time-varying interactions by design
 \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}
With our hyperparameters options and model design, DSM did not
converge in the complex simulation. Due to this issue limitation,
we did not include this method. Compared to CBNN, all three neural network
models also have limitations. DeepSurv is a proportional hazards model and does
not estimate the baseline hazard \citep{katzman2018DeepSurv}. DeepHit
requires an alpha hyperparameter, is restricted to a single distribution
for the baseline hazard and models the survival function directly
\citep{lee2018DeepHit}. Though not included, DSM can model a flexible
baseline hazard, however is not meant for time-varying interactions\citep{dsmPaper}.
The alternative neural network methods match on time, while CBNN models time directly.

To assess performance among these models, we use both IPA and
\(C_{IPCW}\) metrics. Concordance-based measures are commonly used in
survival analysis to compare models and we opt to keep them in our
analyses. However, \(C_{IPCW}\) is a non-proper metric and may cause
misspecified models to appear better than they should
\citep{cindexfails2019}. Therefore, we contextualize our \(C_{IPCW}\)
results in relation to IPA, a proper scoring rule. The model rankings
between \(C_{IPCW}\) and IPA differed for all but FLC. If we consider
Figure \ref{fig:megaPlot} A, E and Table \ref{tab:megaTable} A in conjuction
with FIGURE REF POPTIME, We note a universal drop in \(C_{IPCW}\) performance
for all models. We hide the confidence intervals in Figure \ref{fig:megaPlot} E due to visibility but list them in Table \ref{tab:megaTable} A .
aside from CBNN, all models have a wide confidence band at 75\% of follow-up time.
We expect the optimal model to perform best, but that's not the case looking at  \(C_{IPCW}\).
As such, we

With this interpretation of \(C_{IPCW}\) in mind, we assess a
simulations with minimal noise and three case studies.

The complex simulation requires method that can learn both time-varying
interactions and have a flexible baseline hazard. Here, CBNN demonstrates
a distinct advantage over all other methods. Based on our complex simulation results (Figure
\ref{fig:megaPlot} C, D and Table \ref{tab:megaTable} B), CBNN
outperforms the competitors when time-varying interactions and a complex
baseline hazard are present (after 50\% of survival time, Figure REF POPTIME).
This simulation shows how CBNN can perform
under ideal conditions, while the three case studies
serve to assess its performance in realistic conditions.

In the MM and FLC case studies, flexibility in both
interaction modeling and baseline hazard improved CBNN's performance over
the other models, suggesting that this flexibility aids prediction in
both case studies. In the Prostate case study, the linear models outperform
the neural network ones. CBNN and DeepHit alternate their positions depending
on the survival time and DeepSurv maintains last place.
We attribute this to potential overparameterization in the neural network models
as we did not test for a small number of nodes in each hidden layer, even with dropout.
Though the ranking places the linear models above the neural network ones, the overall
performance is comparable aside from DeepSurv, falling within a small range of IPA scores.

@@

J.I. conceived the proof of concept, developed the theoretical formalism, developed the neural network code base, wrote the majority of the manuscript and performed the simulations and analyses. M.T. and S.B. provided key insights into the core sampling technique (casebase). All authors discussed the results and contributed to the final manuscript. M.T. contributed heavily to the casebase sampling section \ref{case-base-sampling}.

@@

\begin{biography}{\includegraphics[width=60pt,height=70pt,draft]{empty}}{\textbf{Jesse Islam} A PhD Candidate in Quantitative life sciences, studying complex relationships in data in a survival context.}
\end{biography}
